---
title: "Homework 1"
author: "Neal Marquez"
date: "September 28, 2018"
output: pdf_document
---

# Homework 1  

```{R loads, message=F,echo=F,warning=F}
set.seed(123)
library(tidyverse)
```

## Question 1  
The fraction of males in a given population is 0.49. A) What is the probability of sampling 22 males in a random sample of size 30? How about sampling 16 males in a random sample of size 30? Calculate your answers by hand. B) Next, verify your results using R's built-in functions for calculating probabilities. C) Finally, write R code to simulate each of these scenarios many times, and report the results of your simulations.  

### A

Using the function below we can calculate the probability of observing $k$ males in a sample of $n$ individuals if the probability of selecting a male in the total population is $p$. In our test case the $n$ is always $30$, $p$ is always $.49$, and $k$ is either $22$ or $16$. The function follows the formula presented below.

```{r byhand}
handBinom <- function(k,n,p){
    factorial(n) / (factorial(k) * factorial(n-k)) * p^k * (1-p)^(n-k)
}
```

$$
\text{Pr}(k|n,p) = \underbrace{{n \choose k}}_{\text{\# of ways to get k}} \times \underbrace{\prod f_{\text{Bern}}(x_{ij}|p)}_{\text{Pr(getting k)}}
$$
$$
\text{Pr}(k|n,p) = \frac{n!}{k!(n-k)!} p^k (1-p)^{n-k}
$$
$$
\text{Pr}(22|30,.49) = \frac{30!}{22!(30-22)!} .49^{22} (1-.49)^{30-22}
$$
$$
\text{Pr}(22|30,.49) \approx
$$

```{r 1a1}
cat(round(handBinom(22, 30, .49), 6))
```

$$
\text{Pr}(16|30,.49) = \frac{30!}{16!(30-16)!} .49^{16} (1-.49)^{30-16} \\
\text{Pr}(16|30,.49) \approx
$$

```{r 1a2}
cat(round(handBinom(16, 30, .49), 6))
```

### B

Using the `R` built in functions we find that we get the same results.

```{r Part1B}
cat(round(dbinom(22, size=30, prob=.49), digits=6))
cat(round(dbinom(16, size=30, prob=.49), digits=6))
```

### C

We can try and simulate the values in a few different ways in order to get at the true underlying probability. One way is to take a set of $n$ simulations from the Binomial distribution each with $30$ flips and indicate each time an observation equaled either 22 or 16 with a 1 in two separate vectors. We then can take the cumulative mean for each of these vectors and see how the means converge to the truth as the sample size increases.

```{r Part1Cstar}
n <- 10000
singleDF <- tibble(out=rbinom(n=n, size=30, prob=.49), sim=1:n) %>%
    mutate(`Pr(x=22)`=cummean(out == 22), `Pr(x=16)`=cummean(out == 16)) %>%
    gather(key="Statistic", value="Probability", -out, -sim)

singleDF %>%
    ggplot(aes(x=sim, y=Probability, group=Statistic, color=Statistic)) +
    geom_line() +
    theme_classic() +
    geom_hline(yintercept=dbinom(22, size=30, prob=.49), linetype=3) +
    geom_hline(yintercept=dbinom(16, size=30, prob=.49), linetype=3) +
    labs(x="Simulation") +
    ggtitle("Cumulative Estimate of Probability of Outcome")
```

Notice that for the higher probability result we have more variation in the estimate across simulations which is a consequence of the binomial distributions variance. We can also estimate the true mean by taking $n$ simulations where we take $m$ draws from the binomial distribution of $k$ trails and calculate the proportion of values that are equal to either 16 or 22. In the end we are left with a set of estimates that are approximately normally distributed around the true value.

```{r Part1C}
m <- 1000 # Number of simulations (for plotting distribution)
n <- 10000 # number of replications (for calculating individual probs)
tibble(
    Probability=t(sapply(1:m, function(x){
        # Replicate binomial distribution n times with size k and prob .49
        sims <- rbinom(n=n, size=30, prob=.49)
        # calculate probabilty of being  
        c(sum(sims == 22) / n, sum(sims == 16) / n)})) %>%
        c,
    Count=rep(c("22", "16"), each=m)) %>%
    ggplot(aes(x=Probability, fill=Count)) +
    geom_density() +
    facet_wrap(~Count, scales="free") +
    theme_classic() +
    labs(y="Desnity", title="Probabilities of Counts for 1000 Simulations")
```

## Question 2

Consider two random variables X and Y. X can take values equal to 1, 2, or 3, and Y can take values equal to 1, 2, 3, 4. We know the following joint probabilities:

$f_{X,Y}(1,1)=.05$ $f_{X,Y}(2,1)=.07$ $f_{X,Y}(3,1)=.06$  
$f_{X,Y}(1,2)=.08$ $f_{X,Y}(2,2)=\text{???}$ $f_{X,Y}(3,2)=.07$  
$f_{X,Y}(1,3)=.13$ $f_{X,Y}(2,3)=.16$ $f_{X,Y}(3,3)=.15$  
$f_{X,Y}(1,4)=.03$ $f_{X,Y}(2,4)=.04$ $f_{X,Y}(3,4)=.06$  

Using what you know about the relationships between marginal, conditional, and joint probability functions and the information above, compute the following quantities by hand (you should check your results in `R`):

We will place the results of the probabilities presented in a matrix. The last missing cell can be calculated by taking the some of the other probabilities and subtracting that value from 1.

```{R}
probs <- c(.05, .08, .13, .03, .07, NA, .16, .04, .06, .07, .15, .06) %>%
    matrix(4, 3, dimnames=list(paste0("y=", 1:4), paste0("x=", 1:3)))

probs["y=2","x=2"] <- 1 - sum(probs, na.rm=T)

probs
```

Calculations by hand are done by taking the appropriate cells that relate to certain outcomes but we may also use `R` in order to easily calculate marginal probabilities by taking row sums and column sums. A check at the end shows us that all the results are the same.

```{R}
## Hand calculations
(hcalc <- c(
    a=.05 + .08 + .13 + .03,
    b=.07 + .10 + .16 + .04,
    c=.06 + .07 + .15 + .06,
    d=.05 + .07 + .06,
    e=.08 + .10 + .07,
    f=.13 + .16 + .15,
    g=.03 + .04 + .06,
    h=.1,
    i=.08 / sum(.08 + .10 + .07),
    j=.06 / sum(.03 + .04 + .06),
    k=.06 / sum(.06 + .07 + .15 + .06),
    l=.10 / sum(.07 + .10 + .16 + .04)
))
```
```{R}
## R calculations
(rcalc <- c(
    a=sum(probs[,"x=1"]),
    b=sum(probs[,"x=2"]),
    c=sum(probs[,"x=3"]),
    d=sum(probs["y=1",]),
    e=sum(probs["y=2",]),
    f=sum(probs["y=3",]),
    g=sum(probs["y=4",]),
    h=.1,
    i=probs["y=2","x=1"] / sum(probs["y=2",]),
    j=probs["y=4","x=3"] / sum(probs["y=4",]),
    k=probs["y=4","x=3"] / sum(probs[,"x=3"]),
    l=probs["y=2","x=2"] / sum(probs[,"x=2"])
))

cat(all.equal(hcalc, rcalc))

```
