---
title: "Untitled"
author: "Neal Marquez"
date: "November 5, 2018"
output: pdf_document
---

# A
Use an ordered probit model to explain respondents approval of Bush. Using the `llk.oprobit()` function provided in class and `optim()`, fit the model to the variable _bushapp_ with _milforce_, _rbdist_, _econ_, _partyid_, and _yrsofed_ as the only covariates. Report the estimated parameters, their standard errors, and the value of the log likelihood at its maximum.

```{r setup, message=FALSE, warning=FALSE}
rm(list=ls())
library(MASS)
library(tidyverse)
library(mvtnorm)
library(latex2exp)

# Load in the anes data set
anesDF <- "https://faculty.washington.edu/cadolph/mle/nes92con.csv" %>%
    read_csv

# rewrite llk.oprobit4 to be more flexible
llk.oprobit.new <- function(param, formula, data){
    x <- model.matrix(formula[-2], data=data)
    y <- unlist(data[,all.vars(formula)[1]])
    yVals <- sort(unique(y))
    yMat <- sapply(yVals, function(xi) xi == y)
    b <- param[1:ncol(x)]
    xb <- x%*%b
    taus <- c(0, param[(ncol(x)+1):length(param)])
    pMatrix <- matrix(0, nrow=nrow(x), ncol=length(taus) + 1)
    i <- 1
    pMatrix[,i] <- log(pnorm(-xb))
    for(t in taus[2:length(taus)]){
        i <- i + 1
        tlag <- taus[i-1]
        if(t <= tlag){
            pMatrix[,i] <- -(abs(t)*10000)
        }
        else{
            pMatrix[,i] <- log(pnorm(t-xb)-pnorm(tlag-xb))
        }
    }
    pMatrix[,i+1] <- log(1-pnorm(t-xb))

    -sum(pMatrix * yMat)
}




# wrap the fit function of oprobit into a convenient wrapper
oprobit.fit <- function(formula, data){
    xN <- ncol(model.matrix(formula[-2], data=data))
    yN <- length(unique(unlist(data[,all.vars(formula)[1]])))
    tauStarts <- 1:(yN-2)
    names(tauStarts) <- paste0("tau", 2:(yN-1))

    pStart <- c(coef(lm(formula, data=data)), tauStarts)
    optFit <- optim(
        pStart,
        llk.oprobit.new,
        method = "BFGS",
        formula = formula,
        data = data,
        hessian = TRUE)
    optFit$formula <- formula
    optFit
}

fog <- bushapp ~ milforce + rbdist + econ + partyid + yrsofed
modelFit <- oprobit.fit(fog, na.omit(anesDF[,all.vars(fog)]))
cat(paste0("Log Likelihood at the maximum: ", round(-modelFit$value, 4)))
tibble(
    Parameter = names(modelFit$par),
    Estimate = modelFit$par,
    std.err = diag(solve(modelFit$hessian))
)
```

# B
Calculate the probabilities that a respondent who identified themselves as a strong Democrat ( _partyid_ = -3) said they "Strongly Disapproved", "Disapproved", "Approved", or "Strongly Approved" of President Bush given their support or opposition to military force ( _milforce_ = $\{1, 2, 3, 4, 5\}$), holding the remaining covariates at their mean values.  Calculate the probabilities again over the same range of responses for _milforce_ for a respondent who identified themselves as a strong Republican ( _partyid_ = 3). You should have two sets of 20 probabilities, or 40 probabilities total.

```{r}
# Predict function for ordered probit model
modelPredict <- function(model, data, draws=NULL){
    x <- model.matrix(model$formula[-2], data=data)
    yN <- length(model$par) - ncol(x) + 2
    if(is.null(draws)){
        draws_ <- 1
        paramDraws <- matrix(model$par, nrow=draws_, ncol=length(model$par))
        xb <- x %*% paramDraws[,1:ncol(x)]
        tdraws <- matrix(
            c(0, paramDraws[,(ncol(x) + 1):length(model$par)]), ncol=1)
    }
    else{
        draws_ <- draws
        paramDraws <- rmvnorm(draws, model$par, solve(model$hessian))
        xb <- x %*% t(paramDraws[,1:ncol(x)])
        tdraws <- t(cbind(0, paramDraws[,(ncol(x) + 1):length(model$par)]))
    }
    pArray <- array(0, dim=c(nrow(x), yN, draws_))
    for(d in 1:draws_){
        taus <- tdraws[,d]
        i <- 1
        pArray[,i,d] <- pnorm(-xb[,d])
        for(t in taus[2:length(taus)]){
            i <- i + 1
            tlag <- taus[i-1]
            pArray[,i,d] <- pnorm(t-xb[,d])-pnorm(tlag-xb[,d])
        }
        pArray[,i+1,d] <- 1-pnorm(t-xb[,d])
    }
    if(is.null(draws)){
        pArray <- pArray[,,1]
    }
    pArray
}

na.omit(anesDF[,all.vars(fog)]) %>%
    select(-bushapp, -milforce, -partyid) %>%
    summarise_all(mean) %>%
    cbind(expand.grid(milforce=1:5, partyid=c(-3, 3))) %>%
    cbind(round(modelPredict(modelFit, .), 4)) %>%
    select(milforce:`4`)
```

# C

Present the expected probabilities for each category you calculated in part b in a way that would not require your reader to know anything about ordered probit models. Be sure to present confidence intervals, in a fashion of your choosing.

```{r}
nDraws <- 1000
dataMat <- na.omit(anesDF[,all.vars(fog)]) %>%
    select(-bushapp, -milforce, -partyid) %>%
    summarise_all(mean) %>%
    cbind(expand.grid(milforce=1:5, partyid=c(-3, 3)))

simArray <- modelPredict(modelFit, dataMat, draws = nDraws)

appGroups <- c("Strong Disapprove", "Disapprove", "Approve", "Strong Approve")

bind_rows(lapply(1:nDraws, function(d){
    bind_rows(lapply(1:4, function(x) mutate(dataMat, bushapp=x))) %>%
        mutate(draw=d)})) %>%
    mutate(value=c(simArray)) %>%
    group_by(milforce, partyid, bushapp) %>%
    summarize(
        prob = mean(value),
        plwr = quantile(value, probs=.025),
        pupr = quantile(value, probs=.975)) %>%
    mutate(bushapp=factor(appGroups[bushapp], levels=appGroups)) %>%
    mutate(Politics=if_else(
        partyid == 3,
        "Strongly Republican",
        "Strongly Democrat")) %>%
    ggplot(aes(
        x=milforce, y=prob, group=bushapp, fill=bushapp, color=bushapp,
        ymin=plwr, ymax=pupr)) +
    geom_line() +
    geom_ribbon(alpha=.2) +
    theme_classic() +
    facet_wrap(~Politics) +
    labs(color="Bush\nApproval", fill="Bush\nApproval",
         x="Opposition to Military Force", y="Probability") +
    scale_color_brewer(palette="RdYlGn") +
    scale_fill_brewer(palette="RdYlGn")
```

# D

In this part, you are asked to summarize the model estimated in part a. using just two simulated first differences. You may not alter the model itself to eliminate or collapse categories, as this would throw away useful information. __i.__ Simulate the increase in probability of either approving or strongly approving of Bush given a shift from strong opposition to the use of force to strong support for a respondent who is a strong Democrat and otherwise average. Provide a 95% confidence interval. __ii.__ Simulate the increase in probability of either approving or strongly approving of Bush given a shift from strong opposition to the use of force to strong support for a respondent who is a strong Republican and otherwise average. Provide a 95% confidence interval.

```{r}
bind_rows(lapply(1:nDraws, function(d){
    bind_rows(lapply(1:4, function(x) mutate(dataMat, bushapp=x))) %>%
        mutate(draw=d)})) %>%
    mutate(value=c(simArray)) %>%
    mutate(hiApp=bushapp >= 3) %>%
    filter(milforce %in% c(1,5) & hiApp) %>%
    group_by(partyid, draw, milforce) %>%
    summarize(value=sum(value)) %>%
    summarize(deltap=nth(value, 1) - nth(value, 2)) %>%
    group_by(partyid) %>%
    summarize(
        dprob = mean(deltap),
        dplwr = quantile(deltap, probs=.025),
        dpupr = quantile(deltap, probs=.975)) %>%
    mutate(Politics=if_else(
        partyid == 3,
        "Strongly Republican",
        "Strongly Democrat")) %>%
    mutate(Politics=factor(
        Politics,
        levels = c("Strongly Republican", "Strongly Democrat"))) %>%
    ggplot(aes(x=Politics, y=dprob, ymin=dplwr, ymax=dpupr, color=Politics)) +
    geom_point() +
    geom_errorbar(width=.2) +
    theme_classic() +
    labs(y="Change in Probability", x="") +
    guides(color=FALSE) +
    ggtitle(paste(
        "Increase in Approval and Strong Approval Probability",
        "of Bush with Change In Approval of Military Force", sep="\n"))
```

# E
Based upon your work above, briefly discuss of the results of the model.

The model shows that as opposition to military force increases Bush's approval rating drops. This is different depending on political alignment where we can see that on average across all levels of military support Bush's approval rating is lower for strongly Democrats(partyid==-3) than it is for strong Republicans(partyid==3) as seen in section `C`. We also see that as we change support for military force from strongly opposed to strongly in support of, there is a differential effect on the increase of being either approval or strong approval depending on party alignment as seen in section `D`. This effect is probably not signficant, we did not explicitly test that, because there is a large overlap between the two confidence intervals. This is not to say that the effect is the same as we see in the first plot as opposition to military force moves from 5 to 1 most of the weight of the distribution moves to disapprove and approve for strong democrats while for republicans it moves largely to  approve and strongly approve.

# F
The `llk.oprobit()` function is written for ordered data with four categories. Alter the function so that it can be used to estimate a three category ordered probit. Using `R`, create a new variable _econ3_ which collapses _econ_ into a three-category variable. 1 Fit a model to the variable econ3 with _partyid_ as the only covariates to see the effect of an individual's party id on their perception of the economy.  Simply run the model and report the estimated parameters. Perform this fit using your altered function and `optim()`. Bonus [+10 points]: Modify the likelihood function so that it works generically for any number of categories.

Modified function for working with any number of groups is above in the codeblock in section `A`.

```{r, warning=FALSE, message=FALSE}
econDF <- anesDF %>%
    mutate(econ3=as.factor(c(1, 1, 2, 3, 3)[econ])) %>%
    select(econ3, partyid) %>%
    na.omit

eForm <- econ3 ~ partyid
modelEconFit <- oprobit.fit(eForm, econDF)
tibble(
    Parameter = names(modelEconFit$par),
    Estimate = modelEconFit$par,
    std.err = diag(solve(modelEconFit$hessian))
)
```

# G
After you have completed the above parts of the problem, look carefully at the example given in the help for `oprobitsimev` in the `simcf` package. Re-estimate your model using the `polr` function from the `MASS` library. `polr` uses an alternative parameterization of the ordered probit which drops the model's constant in favor of an additional cutpoint (i.e., instead of using a constant and k-2 cutpoints for a k category response variable, `polr` uses k-1 cutpoints and no constant). Note the changes in how we use `oprobitsimev` when using `polr` to estimate ordered probit. Then show that whether we use `optim` or `polr`, we end up with exactly the same predicted probabilities for each category of our response for any given counterfactual scenario.

No matter how we parameterize the model, either with k-2 cut points and an intercept as in the `oprobit` functions or with `polr` and k-1 cutpoints we see the same outcome in both predicted probabilities and counterfactual scenarios. The first plot below shows the predicted probabilities of being in each `econ3` group, that is perceiving the economy as worse, same, or better than before, as a function of party alignment.


```{r, warning=FALSE, message=FALSE}
polrPredict <- function(model, data, draws=NULL){
    X <- model.matrix(update.formula(model$formula[-2], ~ . + 0), data=data)
    if(is.null(draws)){
        pArray <- t(sapply(c(X %*% model$coefficients), function(eta){
            diff(c(0,arm::invlogit(model$zeta - eta), 1))}))
    }
    else{
        pArray <- array(0, dim=c(nrow(X), length(model$zeta) + 1, draws))
        for(d in 1:draws){
            parMeans <- c(model$coefficients, model$zeta)
            bDraws <- rmvnorm(draws, parMeans, vcov(model))
            m_ <- list(
                formula = model$formula,
                coefficients = bDraws[d,1:ncol(X)],
                zeta = bDraws[d,(ncol(X)+1):ncol(bDraws)]
            )
            pArray[,,d] <- polrPredict(m_, data, draws=NULL)
        }
    }
    pArray
}

pModel <- polr(eForm, data=econDF, Hess=TRUE)
pModel$formula <- eForm

pEconDF <- unique(select(econDF, partyid))

econDrawDF <- bind_rows(
    bind_rows(lapply(1:nDraws, function(d){
        bind_rows(lapply(1:3, function(x) mutate(pEconDF, econ3=x))) %>%
            mutate(draw=d)})) %>%
        mutate(value=c(modelPredict(modelEconFit, pEconDF, draws=nDraws))) %>%
        mutate(Model="Optim"),
    bind_rows(lapply(1:nDraws, function(d){
        bind_rows(lapply(1:3, function(x) mutate(pEconDF, econ3=x))) %>%
            mutate(draw=d)})) %>%
        mutate(value=c(polrPredict(pModel, pEconDF, draws=nDraws))) %>%
        mutate(Model="Polr"))

econDrawDF %>%
    group_by(partyid, econ3, Model) %>%
    summarize(
        prob = mean(value),
        plwr = quantile(value, probs=.025),
        pupr = quantile(value, probs=.975)) %>%
    ungroup %>%
    mutate(econ3=c("Better", "Same", "Worse")[econ3]) %>%
    ggplot(aes(
        x=partyid, y=prob, group=econ3, fill=econ3, color=econ3,
        ymin=plwr, ymax=pupr)) +
    geom_line() +
    geom_ribbon(alpha=.2) +
    theme_classic() +
    facet_wrap(~Model) +
    labs(color="Economy", fill="Economy",
         x="Party Identification", y="Probability") +
    scale_x_continuous(
        breaks=c(-2.4,0,2.4),
        labels=c("Strong\nDemocrat", "Independent", "Strong\nRepublican")) +
    ggtitle("Perception of Economy Change Across Party Identification")

econDrawDF %>%
    filter(econ3 != 3) %>%
    group_by(partyid, Model, draw) %>%
    summarize(value=sum(value)) %>%
    summarize(
        dprob = mean(value),
        dplwr = quantile(value, probs=.025),
        dpupr = quantile(value, probs=.975)) %>%
    ggplot(aes(x=partyid, y=dprob, ymax=dpupr, ymin=dplwr)) +
    geom_line() +
    geom_ribbon(alpha=.2) +
    theme_classic() +
    facet_wrap(~Model) +
    labs(x="Party Identification", y=TeX("Pr(Econ $\\neq$ Worse)")) +
    scale_x_continuous(
        breaks=c(-2.7,0,2.7),
        labels=c("Strong\nDemocrat", "Independent", "Strong\nRepublican")) +
    ggtitle(
        "Probability of Economy Not Being Percieved than Worse as Last Year")
```
